---
title: "Chapter6"
author: "BB"
date: "9 december 2018"
output: html_document
---

# Chapter 6
## RATS

First we read the long form of the RATS dataset created in the data wrangling. We factor the categorical variables again, because R does not recognize them as such.

We first make a graphical display of  the weight of all individual rats over time, separated by the different diets (Groups) the rats were exposed to. The rats in group 1 have weight all below 300 grams over time. The rats with diet 2 and 3 vary approximately between 400 and 600 grams. All rats' weight seems to have a tendency to increase over time.
```{r}
# set working directory
setwd("C:/Users/Bram/Desktop/Open data science course/GitHub/IODS-project/data")

#read the long forms of the RATS dataset
RATSL <- read.table(file = "RATSL.txt", header = TRUE, sep = " ", dec = ".")

# Factor variables ID and Group
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

#Access the package ggplot2
library(ggplot2)
library(tidyr)
library(dplyr)

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line() + facet_grid(. ~ Group) + scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) + theme(legend.position = "none") 
```


The rats who have higher weight at the beginning seem to have higher weight at the ending. This phenomenon is known as tracking and can be visualized even more strongly through standardizing the data and making another plot. 
```{r}
# Standardize the weight variable
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)

# Plot again with the standardised rat weight
ggplot(RATSL, aes(x = Time, y = stdWeight, group = ID)) +
  geom_line() + facet_grid(. ~ Group, labeller = labeller(.rows = label_both, .cols = label_value)) + scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) + theme(legend.position = "none") + scale_y_continuous(name = "standardized weight")
```

Because plots with individual data may be slightly chaotic (especially in large datasets), it is useful to make a summary plot and get a general overview of the weight progression over time in the rats with different diets. First we show the mean progression in a line diagram (lines from lowest to highest in order of group 1,2,3).

Then we generate a boxplot leaving out the baseline data (time after timepoint 1). the mean group weight is exposed with the orange diamond. There seem to be no outliers, as there are no data points outside the range of any of the boxes. Hence, we can keep on using the dataset as it is for the following t-test and anova.

```{r}
# All number of days defined as n
n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error of weight by Group (diet) and time in days 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, group = Group)) +
  geom_line() +
  geom_point(size=2) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")

# Create a summary data by Group and Time with mean weight as the summary variable (ignoring baseline week 0).
RATSLS <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, Time) %>%
  summarise(mean=mean(Weight)) %>%
  ungroup()

str(RATSLS$Time)

# Glimpse the data
glimpse(RATSLS)

# Draw a boxplot of the mean versus treatment
ggplot(RATSLS, aes(x = Group, y = mean, group = Group)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=5, fill = "orange") +
  scale_y_continuous(name = "mean Weight over time")
```


I had a hard time finding out how to do the t-test in this data set. Also, I tried to mutate the following way, but it didnt work: 
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", sep  ="\t", header = T). 

RATSLS2 <- RATSLS %>% mutate(baseline = RATS$WD1)



The ANOVA shows there is a significant difference between the groups. However, I suspect this may be a baseline difference and could've been corrected for if the analysis of covariance had been performed properly with the correct baseline variable.
```{r}
#How to perform the t-test in this data set?

# Add the baseline from the original data as a new variable to the summary data. 
RATSLS2 <- RATSLS %>%
  mutate(baseline = RATSL$Time[1])

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSLS2)

# Compute the analysis of variance table for the fitted model
anova(fit)
```

## BPRS

First we ignore the longitudinal structure of repeated measures in this dataset. This gives us a quick general overview of brief psychiatric rating scale scores for individual in two treatment groups. These plots are rather chaotic but show the distribution of individual bprs values. 

```{r}
# set working directory
setwd("C:/Users/Bram/Desktop/Open data science course/GitHub/IODS-project/data")

#read the long forms of the datasets created in data wrangling
BPRSL <- read.table(file = "BPRSL.txt", header = TRUE, sep = " ", dec = ".")

# Factor treatment & subject
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

library(tidyr)
library(dplyr)
library(ggplot2)

#plot BPRSL
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(linetype = subject)) + scale_x_continuous(name = "Week 1-8") + theme(legend.position = "top")

# by treatment separately
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

Then we fit a multiple linear regression model with bprs as target variable and treatment and number of weeks as predictors. It considers that bprs is independent, which is unlikely. This tells us the number of weeks significantly predicts a 2.2 decrease in bprs score per additional week. The treatment has no significant effect.

In the next step we will use a more appropriate 'random intercept model' with the same variables. This model discards the impact of the intercept. It gives variance, effects and associations between predictor and outcome variables. It seems like the variability between subject at the intercept (baseline) is very high, so this needs to be corrected for when statistically comparing these individuals or treatment groups.

```{r}
# create a regression model for bprsl
BPRSL_reg <- lm(bprs ~ treatment + week, BPRSL)

# print out a summary of the model
summary(BPRSL_reg)

# access library lme4
library(lme4)

# Create a random intercept model
BPRSL_reg <- lmer(bprs ~ treatment + week + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRSL_reg)
```

Next, we create a random intercept and random slope model. In this case the intercept and slope may vary between individuals and will be taken into account. Now, the fixed effect correlation shows a stronger association of treatment than the number of weeks with the bprs.

The anova shows that this more accurate 'reg1' model differs significantly from the highly biased 'reg' model. The lower the values of AIC, BIC, logLik, deviance, Chisq, the better the quality of the model.
```{r}
# create a random intercept and random slope model
BPRSL_reg1 <- lmer(bprs ~ treatment + week +  (treatment | subject), data = BPRSL, REML = FALSE)

# print a summary of the model: Every week the bpsr value goes down 2.27 on average. Week and treatment are not correlated
summary(BPRSL_reg1)

# perform an ANOVA test on the two models: the BPRSL_ref1 model differs significantly from the BPRSL_ref model
anova(BPRSL_reg1, BPRSL_reg)
```

Lastly, we add the interaction effects between treatment and number of weeks to The ref2 model. After the addition it still differs significantly from the ref1 model, but differences are minor when comparing to the ref and ref1 model differences.

```{r}
# create a random intercept and random slope model with the interaction
BPRSL_reg2 <- lmer(bprs ~ treatment + week +  (treatment | subject) + treatment * week, data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRSL_reg2)

# perform an ANOVA test on the two models
anova(BPRSL_reg2, BPRSL_reg1)
```


I don't think this is what the plot should look like but this is all I could get. The variability between treatment groups in the first week is high. After that they become fairly equal and change in bprs over time does not differ that much between treatment groups. The fitted values give the lines a coefficient and add a more smooth appearance to the lines in the graph.
```{r}
# draw the plot of BPRSL with the observed bprs values. 
ggplot(BPRSL, aes(x = week, y = bprs, group = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Number of weeks") +
  scale_y_continuous(name = "bprs value") +
  theme(legend.position = "top")

# Create a vector of the fitted values
Fitted <- fitted(BPRSL_reg2)

# Create a new column fitted to RATSL
BPRSL <- mutate(BPRSL, Fitted)

# draw the plot of BPRSL with the Fitted values of weight
ggplot(BPRSL, aes(x = week, y = Fitted, group = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Number of weeks") +
  scale_y_continuous(name = "Fitted bprs value") +
  theme(legend.position = "top")
```

